---
title: "Testing many models with grouped data"
description: |
  Another example of something I do a lot and forget how to do.
author:
  - name: Alice
    url: {}
date: 2022-03-20
output:
  distill::distill_article:
    self_contained: false
    toc: true
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(202203)
```

# Many models!

I often have a situation where I am testing many hypotheses.

How I tested many models in R in the past was to use `lapply` or a loop. 
I don't think there is any problem with that approach, 
I just really like using a pattern these days with grouped data using tidyverse packages.

__The general pattern is__

1. Make the data long (if not already long)
2. Group and nest
3. Mutate to calculate your statistics
4. Un-nest, filter, or select to get your desired output

</br>
There is a great [vignette](https://cran.r-project.org/web/packages/broom/vignettes/broom_and_dplyr.html) on this topic from the {broom} package.

<aside>
Another day, I will add a post on operations on pairwise combinations of 
variables ala the [{corrr} package](https://corrr.tidymodels.org).
</aside>

First, load some packages.

```{r}
library(tidyr, quietly = TRUE) # manipulating data and nesting
library(dplyr, quietly = TRUE) # general data and piping
library(purrr, quietly = TRUE) # i will use purrr::map
library(broom)                 # very good at reformatting model objects
library(palmerpenguins)        # for more fun data
library(survival)              # for time to event models
library(ggplot2)               # to make our plots
theme_set(theme_minimal(base_family = "Avenir")) # for plot appearance
```

## An example: penguin linear models

Let's say we are interested in the association between all the numeric variables
in the {palmerpenguins} penguin dataset and the species. 

If you are not familiar with this dataset, the three penguin species have different features like bill depth, bill length, body mass, and flipper length.

```{r}
head(penguins, 4)
```

Here, we can see that Gentoo are some big penguins and that Adelie has shorter
bill length.

```{r code_folding=TRUE}
penguins %>% 
  pivot_longer(cols = where(is.numeric)) %>% 
  mutate(name = stringr::str_replace_all(name, "_", " "),
         name = stringr::str_wrap(name, width = 10)) %>%
  ggplot(aes(x = species, y = value, fill = species)) + 
  geom_boxplot() + 
  scale_fill_manual(values = c("#0E89BA","#85BAA1","#C16E70")) +
  facet_wrap(~name, scales = "free", nrow = 1) + 
  labs(title = "The penguin species are different",
       x = NULL) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

As a reminder, I will follow the same general pattern above

1. Make the data long - *pivot all the numeric variables* 
2. Group and nest - *group by the variable name*
3. Mutate to calculate your statistics - *variable ~ species*
4. Un-nest, filter, or select to get your desired output 

</br>
Below, I calculated p-values and R-squared values using

1. `stats::lm()`
1. `stats::anova()`
1. `broom::tidy()`

<aside>
Note that you could also use `broom::glance()` to get R-squared
</aside>

```{r}
penguins %>% 
  # tidyr functions to select all the numeric columns and 
  # create a `name` and `value` column
  pivot_longer(cols = where(is.numeric)) %>% 
  group_by(name) %>% 
  # tidyr::nest to create a data frame where each level of the 
  # grouped variable has a single row, and all the other
  # rows and columns are now in a single nested column, `data`
  nest() %>% 
  # use purrr::map to create new nested columns with the objects
  # returned from `lm`, `anova`, `broom::tidy`
  mutate(lm_fit = map(data, 
                      ~ lm(value ~ species, data = .x)),
         r2 = map_dbl(lm_fit, ~summary(.x)$r.squared),
         anova = map(lm_fit, anova),
         tidied = map(anova, tidy)) %>% 
  unnest(tidied) %>%
  # this filter removes the rows with "Residuals"
  filter(term == "species") %>%
  select(-data, -lm_fit, -anova) %>% 
  knitr::kable(digits = 3)
```

We could also do this with `group_modify` in dplyr.

From the documentation:

> group_map(), group_modify() and group_walk() are purrr-style functions
> that can be used to iterate on grouped tibbles.

```{r}
penguins %>% 
  pivot_longer(cols = where(is.numeric)) %>% 
  group_by(name) %>% 
  # there is a litle extra work here to return r.squared
  # group_modify needs the returned value to be a data.frame!
  # so you need to create one
  group_modify( ~tidy(anova(lm(value ~ species, data = .))),
                ~tibble(summary(lm(value ~ species, data = .))$r.squared,
                        .name_repair = ~ c("r2")) ) %>%
  # this filter removes the rows with "Residuals"
  filter(term == "species") %>%
  knitr::kable(digits = 3)
```

## Another example: survival models

I often work with time to event models (survival models). You can also follow this same pattern.

Take for example the `survival::lung` dataset that has some variables
like age, sex, performance status (ECOG and Karnofsky), etc.

```{r}
glimpse(lung)
```

We can repeat the same pattern to test cox proportional hazards
models for these variables individually in univariate models.

```{r}
lung %>% 
  # tidyr to make the data long
  pivot_longer(cols = -c(status, time)) %>% 
  group_by(name) %>% 
  # group the data
  nest() %>% 
  # use purrr::map to create new nested columns with the objects
  mutate(cox_fit = map(data, 
                      ~ coxph(Surv(time, status) ~ value, data = .x)),
         tidied = map(cox_fit, tidy, conf.int = TRUE)) %>% 
  unnest(tidied) %>% 
  select(-data, -cox_fit) %>% 
  knitr::kable(digits = 3)
```

<aside>
You could easily add a `mutate()` here to calculate p-values adjusted
for multiple comparisons.
</aside>

### sessionInfo {.appendix}

```{r}
pander::pander(sessionInfo())
```
