---
title: "Stacking vectors"
description: |
  Don't forget to use stack().
author:
  - name: Alice
    url: {}
date: 2022-03-13
output:
  distill::distill_article:
    self_contained: false
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro
I have recently found a couple of great use cases for the 
`stack()` function from `{utils}`.

Because I want to remind my future self about this, I thought it would make a good short post to test this `{distill}` site that I just created!

# Documentation

From the stack() function documentation:

> "Stacking vectors concatenates multiple vectors into a 
> single vector along with a factor indicating where each 
> observation originated. Unstacking reverses this operation."

# An example

Sometimes, I get a bunch of vectors. Maybe I had multiple files or outputs with various items in them that correspond to different groups. Often, I need to combine these and then check how many of the items exist across multiple groups.

For the purpose of illustration, here I will pretend that I read into R a set of gene names as a named list.

```{r}
my_list <- list(test1 = c("KRAS","EGFR","ERBB2"),
                test2 = c("ERBB2","ERBB3","SPRY2","AR"),
                test3 = c("APC","BRAF"))
```

`stack()` makes a nice tidy data.frame! (*Note that this would also work if the input was a nested list of lists.*)

```{r}
stack(my_list)
```

If you `table()` the result from `stack()`, now you have a nice matrix of the values in each group.

```{r}
table(stack(my_list))
```

The resulting object is a table. You can convert it to a data.frame.

```{r}
as.data.frame.array(table(stack(my_list)))
```

You can also convert the binary matrix to logical (TRUE/FALSE).

```{r}
table(stack(my_list)) > 0
```

Now, imagine a case where you have the table and some values are greater than 1 (because they appeared in a list more than once). You can use a trick to convert to logical and back to numeric 0/1.

```{r}
my_list_w_repeats <- list(
  test1 = c("KRAS","EGFR","ERBB2"),
  test2 = c("ERBB2","ERBB3","SPRY2","AR"),
  test3 = c("APC","APC","APC","BRAF")) # APC is here 3 times

table(stack(my_list_w_repeats))

+(table(stack(my_list_w_repeats)) > 0)
```

# Summary
I forget about this function every once in a while and it is really useful. I also have a [gist](https://gist.github.com/awalsh17/211c9fde41b0d2411a1d5fb61fd78e40) about this.

For fun, here is one way to do this with {dplyr} and {tidyr}. I would like to hear about other ways because I don't find this as intuitive.

```{r}
library(dplyr, quietly = TRUE)

lapply(my_list, function(x) data.frame(genes = x)) %>% 
  bind_rows(.id = "names")
```

Now to make the binary matrix.

```{r}
lapply(my_list, function(x) data.frame(genes = x)) %>% 
  bind_rows(.id = "names") %>%
  count(names, genes) %>%
  tidyr::pivot_wider(names_from = "names",
                     values_from = "n",
                     values_fill = 0)
```

## sessionInfo {.appendix}
```{r}
sessionInfo()
```
